Multilevel constraint decomposition methods \\ for nonlinear variational inequalities

### 2022-08-11 paper/mcd2.tex
done with edit sweep, so box constraints mostly documented

We will approximately solve VI \eqref{eq:vi} for certain nonlinear obstacle problems, over two-dimensional computational domains $\Omega \subset \RR^2$, and the Sobolev space $\mathcal{V}=W_0^{1,p}(\Omega)$, using piecewise-linear finite element (FE) spaces over triangular meshes.  The targeted mesh will be the finest level in a hierarchy of meshes constructed by refinement of a coarsest mesh.  In this Section we construct the corresponding multilevel CD.

@@

Next we define ``box constraints'' \cite{BensonMunson2006}.  Let $\tilde{\mathcal{V}}^j$ denote the set of functions on the $\mathcal{T}^j$ nodes with values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  At the finest-level, suppose that the \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad -\infty < \overline{\gamma}^J, \qquad \text{and} \qquad \underline{\gamma}^J|_{\partial\Omega} \le 0 \le \overline{\gamma}^J|_{\partial\Omega}. \label{eq:fe:boxconstraintrequirements}
\end{equation}
Where $\underline{\gamma}^J=-\infty$ there will no lower constraint, and similarly where $\overline{\gamma}^J=+\infty$ no upper constraint.  Using nodal monotonicity, define
\begin{equation}
\mathcal{K}^J = \left\{v \in \mathcal{V}^J\,:\,\underline{\gamma}^J \le v \le \overline{\gamma}^J\right\}, \label{eq:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of \emph{admissible} FE functions, a closed and convex subset of $\mathcal{V}^J$.  As $\mathcal{V}^J$ functions never take values $\pm\infty$, inequality in \eqref{eq:fineconstraintset} is strict where $\underline{\gamma}^J, \overline{\gamma}^J = \pm \infty$.

@@

@@

The starting point of a \emph{multilevel constraint decomposition} (MCD) \cite{GraeserKornhuber2009,Tai2003} method is to define constraint subsets, the sets generically denoted $\cK_i$ in sections \ref{sec:vi} and \ref{sec:convergence}, based on monotone restrictions of a defect constraint (see below).  Our multilevel V-cycle algorithm will define two types of constraint subsets, distinct sets for the downward and upward parts of the cycle.  We also diverge from the MCD algorithms by Tai \cite{Tai2003} and Gr\"aser \& Kornhuber \cite[Algorithm 4.7]{GraeserKornhuber2009} by following a nonlinear \emph{full approximation scheme} (FAS) philosophy \cite{BrandtLivne2011}.  When $f$ is linear, as in the implemented examples in \cite{GraeserKornhuber2009,Tai2003}, there is no need to coarsen the solution itself, but in the FAS approach here we coarsen the solution iterate using the injection restriction.  Note that in classical linear multigrid only the residual and correction are coarsened.

To make this precise, suppose $w^J \in \cK^J$ is an admissible fine-level iterate which approximates the solution $u^J$ of VI \eqref{eq:fe:vi}.  From this iterate we define the \emph{defect constraint} \cite{GraeserKornhuber2009}
\begin{equation}
\chi^J = \gamma^J - w^J,  \label{eq:fe:defectconstraintfcn}
\end{equation}
a nonpositive function in $\tilde{\mathcal{V}}^J$.  The motivation here is that a corrected iterate is admissible, $w^J + z^J \ge \gamma^J$, if and only if the perturbation $z^J$ exceeds the defect constraint, $z^J \ge \chi^J$.  Next, for $j=J,J-1,\dots,1$ we apply monotone restriction $\maxR$ to define the \emph{level defect constraints},
\begin{equation}
\chi^{j-1} = \maxR \chi^j. \label{eq:fe:chilevels}
\end{equation}
so that $\chi^j \le \chi^{j-1}$.  A $J=4$ example is shown in Figure \ref{fig:chiphilevels}.

We also denote the differences between the level defect constraints by
\begin{equation}
\phi^j = \chi^j - \chi^{j-1}, \label{eq:fe:philevels}
\end{equation}
and these are nonpositive functions.  (Definition \eqref{eq:fe:philevels} uses $\chi^{-1}=0$.)  Functions $\phi^j$ are the gaps between the $\chi^j$ shown in Figure \ref{fig:chiphilevels}, with $\phi^0=\chi^0$ in particular.  Note the telescoping sum decomposition $\sum_{j=0}^J \phi^j = \chi^J$.

%REGENERATE using https://github.com/bueler/mg-glaciers:
%$ cd mg-glaciers/py/
%$ [modify decomposition() inside visualize.py to turn off axes and print \gamma for obstacle]
%$ ./obstacle.py -J 5 -jcoarse 1 -random -randommodes 8 -diagnostics -o defect.pdf
%fine level 5 (m=63): using 20 V(1,0) cycles -> 38.750 WU
\begin{figure}
\includegraphics[width=0.55\textwidth]{fixfigs/chiphilevels.pdf}
\caption{Defect constraint $\chi^J = \gamma^J - w^J$ is decomposed using monotone restriction.}
\label{fig:chiphilevels}
\end{figure}

For the downward part of the multigrid V-cycle we will use the $\phi^j$ as constraints, and for the upward part the $\chi^j$.  That is, let
\begin{equation}
\mathcal{D}^j = \{v \in \mathcal{V}^j \,:\, v\ge \phi^j\}, \qquad \mathcal{U}^j = \{v \in \mathcal{V}^j \,:\, v\ge \chi^j\} \label{eq:fe:constraintsets}
\end{equation}
define \emph{downward} and \emph{upward constraint subsets} of the FE subspaces $\mathcal{V}^j$, respectively.  The upward sets are larger, $\mathcal{D}^j \subset \mathcal{U}^j$, and set decompositions $\mathcal{U}^j = \sum_{i=0}^j \mathcal{D}^i$ apply in the same sense as \eqref{eq:constraintdecomp}.

Returning to the central MCD idea, we will correct the finest-level iterate $w^J\in \mathcal{K}^J$ by adding modifications from each coarser subspace $\mathcal{V}^j$.  By nesting \eqref{eq:fe:nestedspaces} these corrections live in the fine-level subspace $\mathcal{V}^J$, but the corrected iterate needs to be admissible as well.  The accumulated corrections, from the constraint subsets $\mathcal{D}^j$ and $\mathcal{U}^j$, over the whole multigrid V-cycle, compute a new iterate in $\mathcal{K}^J$.

@@

\begin{figure}[ht]
\begin{center}
\input{tikz/nmcdvcycle.tex}
\end{center}
\caption{An NMCD V-cycle computes downward corrections $y_j \in \mathcal{D}^j$, but the upward corrections $z_j$ expand into larger constraint subsets $\mathcal{U}^j$.}
\label{fig:nmcdvcycle}
\end{figure}

@@

These ideas come together in Algorithm \ref{alg:nmcd}, an NMCD V-cycle.  This procedure acts in-place on the final argument $w^J$, but it leaves the other inputs $\ell^J,\gamma^J$ unchanged.

@@

\appendix
\section{Reductions of the NMCD algorithm}

@@

>>>>

We will approximately solve VI \eqref{eq:vi} for certain nonlinear obstacle problems, over two-dimensional computational domains $\Omega \subset \RR^2$, and the Sobolev space $\mathcal{V}=W_0^{1,p}(\Omega)$, using piecewise-linear finite element (FE) spaces over triangular meshes.  The targeted mesh will be the finest level in a hierarchy of meshes constructed by refinement of a coarsest mesh.  In this Section we construct the corresponding multilevel CD.  Its constraint sets are designed to allow multigrid coarser-level corrections (next Section) in a manner which does not violate the inequality constraints, and which also does not introduce unnecessary high frequencies into the coarser-level problems.

@@

Next we define ``box constraints'' \cite{BensonMunson2006}.  Let $\tilde{\mathcal{V}}^j$ denote the set of functions on the $\mathcal{T}^j$ nodes with values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  At the finest-level, suppose that the \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad -\infty < \overline{\gamma}^J, \qquad \text{and} \qquad \underline{\gamma}^J|_{\partial\Omega} \le 0 \le \overline{\gamma}^J|_{\partial\Omega}. \label{eq:fe:boxconstraintrequirements}
\end{equation}
Where $\underline{\gamma}^J=-\infty$ there will no lower constraint, and similarly where $\overline{\gamma}^J=+\infty$ no upper constraint.  Using nodal monotonicity, define
\begin{equation}
\mathcal{K}^J = \left\{v \in \mathcal{V}^J\,:\,\underline{\gamma}^J \le v \le \overline{\gamma}^J\right\}, \label{eq:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of \emph{admissible} FE functions, a nonempty, closed, and convex subset of $\mathcal{V}^J$.  As $\mathcal{V}^J$ functions never take values $\pm\infty$, inequality in \eqref{eq:fineconstraintset} is strict where $\underline{\gamma}^J, \overline{\gamma}^J = \pm \infty$.

@@

To define the constraint subsets, the sets generically denoted $\cK_i$ in Sections \ref{sec:cd}--\ref{sec:convergence}, in the current multilevel FE context, suppose $w^J \in \cK^J$ is an admissible fine-level iterate which approximates the solution $u^J$ of VI \eqref{eq:fe:vi}.  From this iterate we define the finest-level \emph{lower and upper defect constraints} \cite{GraeserKornhuber2009}
\begin{equation}
\underline{\chi}^J = \underline{\gamma}^J - w^J, \qquad \overline{\chi}^J = \overline{\gamma}^J - w^J, \label{eq:fe:defectconstraints}
\end{equation}
functions in $\tilde{\mathcal{V}}^J$ satisfying $\underline{\chi}^J \le 0 \le \overline{\chi}^J$.  The motivation here is that a corrected iterate is admissible, $\underline{\gamma}^J \le w^J + z^J \le \overline{\gamma}^J$, if and only if the perturbation $z^J$ is between the defect constraints, $\underline{\chi}^J \le z^J \le \overline{\chi}^J$.

For $j=J,J-1,\dots,1$ we apply monotone restrictions to define the \emph{level defect constraints},
\begin{equation}
\underline{\chi}^{j-1} = \maxR \underline{\chi}^j, \qquad \overline{\chi}^{j-1} = \minR \overline{\chi}^j. \label{eq:fe:chilevels}
\end{equation}
Observe that
\begin{equation}
\underline{\chi}^{J} \le \dots \le \underline{\chi}^0 \le 0 \le \overline{\chi}^0 \le \dots \le \overline{\chi}^J. \label{eq:fe:chiordering}
\end{equation}
A $J=4$ example, with no upper obstacles ($\overline{\chi}^j=+\infty$), is shown in Figure \ref{fig:chiphilevels}.

%REGENERATE using https://github.com/bueler/mg-glaciers:
%$ cd mg-glaciers/py/
%$ [modify decomposition() inside visualize.py to turn off axes and print \gamma for obstacle]
%$ ./obstacle.py -J 5 -jcoarse 1 -random -randommodes 8 -diagnostics -o defect.pdf
%fine level 5 (m=63): using 20 V(1,0) cycles -> 38.750 WU
\begin{figure}
\includegraphics[width=0.65\textwidth]{fixfigs/chiphilevels.pdf}
\caption{The lower defect constraint $\underline{\chi}^J = \underline{\gamma}^J - w^J$ is decomposed using maximum restriction: $\underline{\chi}^{j-1} = \maxR \underline{\chi}^j$.}
\label{fig:chiphilevels}
\end{figure}

We also compute the differences between the level defect constraints:
\begin{equation}
\underline{\phi}^j = \underline{\chi}^j - \underline{\chi}^{j-1}, \qquad \overline{\phi}^j = \overline{\chi}^j - \overline{\chi}^{j-1},  \label{eq:fe:philevels}
\end{equation}
using $\underline{\chi}^{-1}=\overline{\chi}^{-1}=0$.  Again these constraints bracket zero,
\begin{equation}
\underline{\phi}^j \le 0 \le \overline{\phi}^j,  \label{eq:fe:phibrackets}
\end{equation}
but these functions are not ordered as in \eqref{eq:fe:chiordering}, but telescoping sums hold for $j=0,1,\dots,J$:
\begin{equation}
\sum_{i=0}^j \underline{\phi}^i = \underline{\chi}^j, \qquad \sum_{i=0}^j \overline{\phi}^i = \overline{\chi}^j.  \label{eq:fe:telescoping}
\end{equation}

A multilevel V-cycle is defined in the next Section.  For the downward direction we will use the differences $\underline{\phi}^j,\overline{\phi}^j$ as constraints, and for the upward part the defect constraints $\underline{\chi}^j,\overline{\chi}^j$.  That is,
\begin{align}
\mathcal{D}^j = \{v \in \mathcal{V}^j \,:\, \underline{\phi}^j \le v \le \overline{\phi}^j\}, \label{eq:fe:constraintsets} \\
\mathcal{U}^j = \{v \in \mathcal{V}^j \,:\, \underline{\chi}^j \le v \le \overline{\chi}^j\}, \notag
\end{align}
are the \emph{downward} and \emph{upward constraint subsets} of the FE subspaces $\mathcal{V}^j$, respectively.  Observe that the upward sets are larger, $\mathcal{D}^j \subset \mathcal{U}^j$, and in fact by \eqref{eq:fe:telescoping} the set decompositions
\begin{equation}
\mathcal{U}^j = \sum_{i=0}^j \mathcal{D}^i \label{eq:fe:constraintdecomp}
\end{equation}
apply in the same sense as \eqref{eq:constraintdecomp}.  Also note that the finest-level upward constraint set is equivalent to the original constraint set $\mathcal{K}^J$; for all $z^J \in \mathcal{V}^J$,
\begin{equation}
z^J \in \mathcal{U}^J \quad \iff \quad w^J+z^J \in \mathcal{K}^J. \label{eq:fe:finestlevelequivalent}
\end{equation}

FIXME what are the CDs?  what are the decomposition operators?

@@

Our \emph{nonlinear multilevel constraint decomposition} (NMCD) V-cycle algorithm in this section will use the two types of constraint subsets in \eqref{eq:fe:constraintsets}, with distinct sets for the downward and upward parts of the cycle.  We also diverge from the multilevel CD algorithms by Tai \cite{Tai2003} and Gr\"aser \& Kornhuber \cite[Algorithm 4.7]{GraeserKornhuber2009} by following a nonlinear \emph{full approximation scheme} (FAS) philosophy \cite{BrandtLivne2011}.  In the FAS approach here we must coarsen the solution iterate itself, using the injection restriction, in constrast to classical linear multigrid, wherein only the residual and correction are coarsened, and in constrast to the implemented examples in \cite{GraeserKornhuber2009,Tai2003} in which $f$ is linear.

FIXME from here

We will correct the finest-level iterate $w^J\in \mathcal{K}^J$ by adding modifications from each coarser subspace $\mathcal{V}^j$.  By nesting \eqref{eq:fe:nestedspaces} these corrections live in the fine-level subspace $\mathcal{V}^J$, but the corrected iterate needs to be admissible as well.  The accumulated corrections, from the constraint subsets $\mathcal{D}^j$ and $\mathcal{U}^j$, over the whole multigrid V-cycle, compute a new iterate in $\mathcal{K}^J$.

@@

\begin{figure}[ht]
\begin{center}
\input{tikz/nmcdvcycle.tex}
\end{center}
\caption{An NMCD V-cycle computes downward corrections $y_j \in \mathcal{D}^j$, but the upward corrections $z_j$ may expand into larger constraint subsets $\mathcal{U}^j$.}
\label{fig:nmcdvcycle}
\end{figure}

@@

These ideas come together in Algorithm \ref{alg:nmcd}, an NMCD V-cycle.  This procedure acts in-place on the final argument $w^J$, but it leaves the other vector inputs $\ell^J,\gamma^J$ unchanged.

@@

\appendix
\section{Reduction of NMCD to unconstrained FAS}

@@

\section{Reduction of NMCD for linear operators}
